Polynomial phase estimation by phase unwrapping

### 2013-02-22 paper.tex
minor

In this paper we consider the estimator that results from unwrapping the phase in a least squares manner.  We call this the \emph{least squares unwrapping} (LSU) estimator~\cite{McKilliam2009asilomar_polyest_lattice, McKilliamFrequencyEstimationByPhaseUnwrapping2009,McKilliam2010thesis}.  It was shown in \cite{McKilliam2009asilomar_polyest_lattice, McKilliamFrequencyEstimationByPhaseUnwrapping2009} that the LSU estimator can be represented as a \emph{nearest lattice point problem}~\cite{Agrell2002}, and Monte-Carlo simulations were used to show the LSU estimator's favourable statistical performance. %Moreover, it was noticed that the LSU appears to work correctly for coefficients anywhere in the identifiable region.  
A drawback of the LSU estimator is that computing a nearest lattice point is, in general, computationally difficult.  In \cite{McKilliam2009asilomar_polyest_lattice}, two standard techniques were considered, the \emph{sphere decoder}~\cite{Pohst_sphere_decoder_1981,Viterbo_sphere_decoder_1999} that exactly computes a nearest lattice point, and \emph{Babai's nearest plane algorithm}~\cite{Babai1986} that only produces an approximation. The sphere decoder was observed to have excellent statistical performance but can only be computed efficiently for small $N$ less than about $60$.  Babai's algorithm is computationally simple, but its statistical performance was observed to be comparatively poor at low signal to noise ratio (SNR). A major point of interest is that the lattices considered are not \emph{random}, and may admit fast nearest point algorithms.  This has been studied in~\cite{McKilliam2010thesis} where polynomial time algorithms were found that compute the nearest point exactly.  Although polynomial time, the algorithms are still computationally demanding in practice.  Nevertheless fast (exact or approximate) algorithms may exist for these special lattices, that are yet to be discovered.  In this paper we apply another general purpose approximate nearest point algorithm, called the $K$-best method~\cite{Zhan2006_K_best_sphere_decoder} for approximating the LSU estimator.  We show that this provides near sphere decoder (exact) performance, but can be computed in a reasonable amount of time if $N$ is not too large (approximately $N$ less than $1000$).  %The $K$-best method is motivated by sequential decoding algorithms and, in the context of polynomial phase estimation, bares some resemblance to the Bayesian unwrapping algorithm studied by Morelande~\cite{Morelande_bayes_unwrapping_2009_tsp}.
%We show by Montecarlo simulation that the LSU estimator approximated using the $K$-best method is statistically superior to state-of-the-art estimators from the literature including the HAF and CPF.  We also compare the computational performance of the estimators and find that BLERG.

In addition to the above computational results a major purpose of this paper is to study the asymptotic properties of the LSU estimator.  Under mild assumptions about the distribution of the noise we prove that the LSU estimator is strongly consistent, that is, the estimated coefficients converge (almost surely) to the true coefficients as the number of observations $N$ grows.  %Similar results were stated without proof in~\cite{McKilliam_polyphase_est_icassp_2011}. 
The proof makes use of an elementary result about the number of arithmetic progressions contained inside subsets of $\{1,2,\dots,N\}$~\cite{Erdos_on_some_sequence_of_integers1936,Szemeredi_setint_no_k_arth1975,Gowers_new_proof2001}.  %This proof technique appears to be novel, and may be useful in other problems.  
%The proof of asymptotic normality is complicated by the fact that the objective function corresponding with the LSU estimator is not differentiable everywhere.  Empirical process techniques~\cite{Pollard_new_ways_clts_1986,Pollard_asymp_empi_proc_1989,van2009empirical,Dudley_unif_central_lim_th_1999} and results from the literature on hyperplane arrangements~\cite{Chazelle_discrepency_method_2000,Matousek_lect_disc_geom_2002} become useful here.  
This proof technique appears to be novel and may be useful for purposes other than polynomial phase estimation, and in particular other applications involving data that is `wrapped' in some sense.  Potential candidates are the phase wrapped images observed in modern radar and medical imaging devices such as synthetic aperture radar and magnetic resonance imaging~\cite{Katkovnik_phasela_2009, Nico_phaseunwrappingSAR_2000,Friedlander_PD_phaseunwrapping_1996}.

The analysis of most existing polynomial phase estimators make what is known as a `small noise' assumption~\cite{Peleg_linear_FM,Oshea_cpf_2004,Djurovic_haf_cpf_2012,Djuric_phase_unwrap_chirp_1990}.  This amounts to assuming that the estimator $\hat{\mu}_k$ of the $k$th coefficient is such that $N^{k}(\tilde{\mu}_k -  \hat{\mu}_k) \rightarrow 0$ (in some sense) as $N \rightarrow \infty$.  This assumption is used to justify the use of first and second order expansions~\cite{vonMises_diff_stats_1947,vanDerVart1971_asymptotic_stats} (typically called a `perturbation analysis' in this literature~\cite{Peleg_linear_FM,Porat_asympt_HAF_DPT_1996,Oshea_cpf_2004}) to obtain expressions for the asymptotic mean square error of the estimator in question.  The `small noise' assumption is usually justified by Monte-Carlo simulations showing the asymptotic mean square error obtained by the perturbation analysis to be accurate provided that the signal to noise ratio (SNR) is larger than a `threshold' value (see the simulations in Section~\ref{sec:simulations} for examples).  Typically one observes the threshold to occur at smaller values of SNR as $N$ increases.  A problem with the `small noise' assumption is that it is not known that the threshold continues to occur at increasingly smaller values of SNR as $N$ increases, or if it instead converges to some finite (but perhaps small) value of SNR.  The consequence of the later occurance being that the estimator will be innaccurate if the SNR is smaller than this finite value, regardless of the number of observations $N$ obtained.  Our analysis of the LSU estimator requires no `small noise' assumpion and guarantees that no such finite value of SNR exists for the LSU estimator.  As a result the LSU estimator will be accurate for signals with arbitrarily small SNR provided that the number of observations is sufficiently large.

@@

A key property of the LSU estimator is that it functions correctly for polynomial phase parameters contained anywhere inside the identifiable region described in Section~\ref{sec:ident_aliasing} and~\cite{Xia_dynamic_range_pps_2001,McKilliam2009IndentifiabliltyAliasingPolyphase}.  This is the maximum possible range.  By constrast popular estimators such as the HAF and CPF operate correctly only for parameters in a range much smaller than the identifiable region.  This property has been studied previously~\cite{Djurovic_aliasing_2012,McKilliam2009IndentifiabliltyAliasingPolyphase,Zhou_lag_diverity_PPT_1997,Xia_dynamic_range_pps_2001} and can be a critical limitation for some applications, for example, interference supression in the pressence of a jammer~\cite{Djukanovic:2011:PMN:1950994.1951252,Djurovic_aliasing_2012}.  Approaches exist that extend the range of parameters suitable for the HAF~\cite{Zhou_lag_diverity_PPT_1997,Djurovic_aliasing_2012}, but the resulting estimators typically have poor statistical performance.  For this reason, the LSU estimator might be the only presently existing estimator that is viable for certain applications, such as interference supression.  The theoretical results given in Section~\ref{sec:least-squar-unwr} prove that the LSU estimator functions correctly over the entire identifiable region.  To our knowledge this is the first time a result of this type has been obtained for any polynomial phase estimator.

The paper is organised in the following way. Section~\ref{sec:lattice-theory} describes some required concepts from lattice theory.  In Section~\ref{sec:ident_aliasing} we describe the identifiable region that was also derived in~\cite{McKilliam2009IndentifiabliltyAliasingPolyphase}.  These identifiability results are required in order to properly understand the statistical properties of the LSU estimator.  Section~\ref{sec:least-squar-unwr} describes the LSU estimator and states a theorem asserting the estimator to be strongly consistent and asymptotically normally distributed under some mild assumptions on the noise $\{X_n\}$. The proof of strong consistency is given in Section~\ref{sec:strongconstproof}.  Space restrictions make it impossible to give a proof of asymptotic normality in this paper.  A complete proof of normality is available in~\cite{McKilliam_arxiv_pps_unwrapping_2012}.  Section~\ref{sec:simulations} describes the results of Monte Carlo simulations that compare the performance of the LSU estimator with existing estimators including the HAF and CPF.  These simulations agree with the derived asymptotic properties. BLERG: benchmarks.

>>>>

In this paper we consider the estimator that results from unwrapping the phase in a least squares manner.  We call this the \emph{least squares unwrapping} (LSU) estimator~\cite{McKilliam2009asilomar_polyest_lattice, McKilliamFrequencyEstimationByPhaseUnwrapping2009,McKilliam2010thesis}.  It was shown in \cite{McKilliam2009asilomar_polyest_lattice, McKilliamFrequencyEstimationByPhaseUnwrapping2009} that the LSU estimator can be represented as a \emph{nearest lattice point problem}~\cite{Agrell2002}, and Monte-Carlo simulations were used to show the LSU estimator's favourable statistical performance. %Moreover, it was noticed that the LSU appears to work correctly for coefficients anywhere in the identifiable region.  
A drawback of the LSU estimator is that computing a nearest lattice point is, in general, computationally difficult.  In \cite{McKilliam2009asilomar_polyest_lattice}, two standard techniques were considered, the \emph{sphere decoder}~\cite{Pohst_sphere_decoder_1981,Viterbo_sphere_decoder_1999,Agrell2002} that exactly computes a nearest lattice point, and \emph{Babai's nearest plane algorithm}~\cite{Babai1986} that only produces an approximation. The sphere decoder was observed to have excellent statistical performance but can only be computed efficiently for small $N$ less than about $60$.  Babai's algorithm is computationally simple, but its statistical performance was observed to be comparatively poor at low signal to noise ratio (SNR). A major point of interest is that the lattices considered has significant structure, and may admit fast nearest point algorithms.  This has been studied in~\cite{McKilliam2010thesis} where polynomial time algorithms were found that compute the nearest point exactly.  Although polynomial time, the algorithms are still computationally demanding in practice.  Nevertheless fast (exact or approximate) algorithms may exist for these special lattices, that are yet to be discovered.  In this paper we apply another general purpose approximate nearest point algorithm, called the $K$-best method~\cite{Zhan2006_K_best_sphere_decoder} for approximating the LSU estimator.  We show that this provides near sphere decoder (exact) performance, but can be computed in a reasonable amount of time if $N$ is not too large (approximately $N$ less than $1000$).  %The $K$-best method is motivated by sequential decoding algorithms and, in the context of polynomial phase estimation, bares some resemblance to the Bayesian unwrapping algorithm studied by Morelande~\cite{Morelande_bayes_unwrapping_2009_tsp}.
%We show by Montecarlo simulation that the LSU estimator approximated using the $K$-best method is statistically superior to state-of-the-art estimators from the literature including the HAF and CPF.  We also compare the computational performance of the estimators and find that BLERG.

In addition to the above computational results a major purpose of this paper is to study the asymptotic properties of the LSU estimator.  Under mild assumptions about the distribution of the noise we prove that the LSU estimator is strongly consistent, that is, the estimated coefficients converge almost surely to the true coefficients as the number of observations $N$ grows.  %Similar results were stated without proof in~\cite{McKilliam_polyphase_est_icassp_2011}. 
The proof makes use of an elementary result about the number of arithmetic progressions contained inside subsets of $\{1,2,\dots,N\}$~\cite{Erdos_on_some_sequence_of_integers1936,Szemeredi_setint_no_k_arth1975,Gowers_new_proof2001}.  %This proof technique appears to be novel, and may be useful in other problems.  
%The proof of asymptotic normality is complicated by the fact that the objective function corresponding with the LSU estimator is not differentiable everywhere.  Empirical process techniques~\cite{Pollard_new_ways_clts_1986,Pollard_asymp_empi_proc_1989,van2009empirical,Dudley_unif_central_lim_th_1999} and results from the literature on hyperplane arrangements~\cite{Chazelle_discrepency_method_2000,Matousek_lect_disc_geom_2002} become useful here.  
This proof technique appears to be novel and may be useful for purposes other than polynomial phase estimation, and in particular other applications involving data that is `wrapped' in some sense.  Potential candidates are the phase wrapped images observed in modern radar and medical imaging devices such as synthetic aperture radar and magnetic resonance imaging~\cite{Katkovnik_phasela_2009, Nico_phaseunwrappingSAR_2000,Friedlander_PD_phaseunwrapping_1996}.

The analysis of most existing polynomial phase estimators make what is known as a `small noise' assumption~\cite{Peleg_linear_FM,Oshea_cpf_2004,Djurovic_haf_cpf_2012,Djuric_phase_unwrap_chirp_1990}.  This amounts to assuming that the estimator $\hat{\mu}_k$ of the $k$th coefficient is such that $N^{k}(\tilde{\mu}_k -  \hat{\mu}_k) \rightarrow 0$ (in some sense) as $N \rightarrow \infty$.  This assumption is used to justify the use of first and second order expansions~\cite{vonMises_diff_stats_1947,vanDerVart1971_asymptotic_stats} (typically called a `perturbation analysis' in this literature~\cite{Peleg_linear_FM,Porat_asympt_HAF_DPT_1996,Oshea_cpf_2004}) to obtain expressions for the asymptotic mean square error of the estimator in question.  The `small noise' assumption is usually justified by Monte-Carlo simulations showing the asymptotic mean square error obtained by the perturbation analysis to be accurate provided that the signal to noise ratio (SNR) is larger than a `threshold' value (see the simulations in Section~\ref{sec:simulations} for examples).  Typically one observes the threshold to occur at smaller values of SNR as $N$ increases.  A problem with the `small noise' assumption is that it is not known that the threshold continues to occur at increasingly smaller values of SNR as $N$ increases, or if it instead converges to some finite (but perhaps small) value of SNR.  The consequence of the later being that the estimator will be innaccurate if the SNR is smaller than this finite value, regardless of the number of observations $N$ obtained.  Our analysis of the LSU estimator requires no `small noise' assumpion and guarantees that no such finite value of SNR exists for the LSU estimator.  As a result the LSU estimator can be accurate for signals with arbitrarily small SNR provided that the number of observations is sufficiently large.

@@

A key property of the LSU estimator is that it functions correctly for polynomial phase parameters contained anywhere inside the identifiable region described in Section~\ref{sec:ident_aliasing} and~\cite{Xia_dynamic_range_pps_2001,McKilliam2009IndentifiabliltyAliasingPolyphase}.  This is the maximum possible range of parameters.  By constrast popular estimators such as the HAF and CPF operate correctly only for parameters in a range much smaller than the identifiable region.  This property has been studied previously~\cite{Djurovic_aliasing_2012,McKilliam2009IndentifiabliltyAliasingPolyphase,Zhou_lag_diverity_PPT_1997,Xia_dynamic_range_pps_2001} and can be a critical limitation for some applications, for example, interference supression in the pressence of a jammer~\cite{Djukanovic:2011:PMN:1950994.1951252,Djurovic_aliasing_2012}.  Approaches exist that extend the range of parameters suitable for the HAF~\cite{Zhou_lag_diverity_PPT_1997,Djurovic_aliasing_2012}, but the resulting estimators typically have poor statistical performance.  For this reason, the LSU estimator may be the only presently existing estimator that is viable for certain applications, such as interference supression.  The theoretical results given in Section~\ref{sec:least-squar-unwr} prove that the LSU estimator functions correctly over the entire identifiable region.  To our knowledge this is the first time a result of this type has been obtained for any polynomial phase estimator.

The paper is organised in the following way. Section~\ref{sec:lattice-theory} describes some required concepts from lattice theory.  In Section~\ref{sec:ident_aliasing} we describe the identifiable region that was also derived in~\cite{McKilliam2009IndentifiabliltyAliasingPolyphase}.  These identifiability results are required in order to properly understand the statistical properties of the LSU estimator.  Section~\ref{sec:least-squar-unwr} describes the LSU estimator and states a theorem asserting the estimator to be strongly consistent and asymptotically normally distributed under some mild assumptions on the noise $\{X_n\}$. The proof of strong consistency is given in Section~\ref{sec:strongconstproof}.  Space restrictions make it impossible to give a proof of asymptotic normality in this paper.  A complete proof of normality is available in~\cite{McKilliam_arxiv_pps_unwrapping_2012}.  Section~\ref{sec:simulations} describes the results of Monte Carlo simulations that compare the LSU estimator with existing estimators including the HAF and CPF.  These simulations agree with the derived asymptotic properties. BLERG: benchmarks.
